Using the internal language of toposes in commutative algebra

### 2017-09-16 slides-pssl101.tex
to do
Slides almost finished

\usepackage{mathtools}
%\newcommand{\vcentcolon}{:}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
\hypersetup{colorlinks=true}

@@

% Orally: Laws of reasoning do not only apply to ordinary mathematical objects.
% They also apply to mathematical objects of different toposes. By cleverly
% choosing those topos, we can pretend that ...
\frame[t]{\frametitle{Quick summary}
  By employing the internal language of toposes in various ways, you can
  pretend that:
  \begin{enumerate}
    \item Sheaves of modules are plain modules.
    \item Schemes are sets: \[ \PP^2_S = \{
    [x_0\mathbin{:}x_1\mathbin{:}x_2] \,|\, x_0 \neq 0 \vee x_1 \neq 0 \vee x_2 \neq
    0 \}. \]
    \item Reduced rings are Noetherian and in fact fields.
  \end{enumerate}
}
% Any theorem automatically yields infinitely new theorems; sometimes, these
% are useful, and sometimes, they are indeed very nice.

@@

\subsection{What is the internal language?}
\frame[t]{\frametitle{What is the internal language?}
  The internal language of a topos~$\E$ allows to
  \begin{enumerate}
    \item construct objects and morphisms of the topos,
    \item formulate statements about them and
    \item prove such statements
  \end{enumerate}
  in a \hil{naive element-based} language:

@@

  The \hil{little Zariski topos} of~$A$ is the category~$\Sh(\Spec(A))$ of
  set-valued sheaves on its spectrum. It contains a ring object~$A^\sim$
  with~$A^\sim(D(f)) = A[f^{-1}]$.

  \begin{block}{Motto}The structure sheaf~$A^\sim$ is a reification of
  all of the stalks~$A_\ppp$.\end{block}

@@

  \hil{Proof.} The structure sheaf~$A^\sim$ is the localisation
  \[ \underline{A}[\F^{-1}] \]
  of the constant sheaf~$\underline{A}$ at the \hil{generic filter}~$\F$, a
  sheaf with
  \[ \F_\ppp = A \setminus \ppp. \]
  The rings~$A$ and~$\underline{A}$ share all first-order
  properties.
\end{frame}

@@

  \hil{Theorem.} There exists $f \neq 0$ in $A$ such that
  \begin{enumerate}
    \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
    \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$.
  \end{enumerate}
  \pause

@@

  \hil{Proof.} Internally in~$\Sh(\Spec(A))$,
  \begin{enumerate}
    \item $B^\sim$ and~$M^\sim$ are \hil{not not} free
    over~$A^\sim$
    \item $A^\sim \to B^\sim$ is \hil{not not} of finite presentation,
    and
    \item $M^\sim$ is \hil{not not} finitely presented as a module
    over~$B^\sim$.
  \end{enumerate}
\end{frame}

@@

>>>>

\usepackage{mathtools}
%\newcommand{\vcentcolon}{:}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns}
\hypersetup{colorlinks=true}

@@

% Orally: Laws of reasoning do not only apply to ordinary mathematical objects.
% They also apply to mathematical objects of different toposes. By cleverly
% choosing those topos, we can pretend that ...
\frame[t]{\frametitle{Quick summary}
  By employing the internal language of toposes in various ways, you can
  pretend that:
  \begin{enumerate}
    \item Sheaves of modules are plain modules.
    \item Schemes are sets: \[ \PP^2_S = \{
    [x_0\mathbin{:}x_1\mathbin{:}x_2] \,|\, x_0 \neq 0 \vee x_1 \neq 0 \vee x_2 \neq
    0 \}. \]
    \item Reduced rings are Noetherian and in fact fields.
  \end{enumerate}

% \scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}
}
% Any theorem automatically yields infinitely new theorems; sometimes, these
% are useful, and sometimes, they are indeed very nice.

@@

\subsection[Kripke--Joyal semantics]{Kripke--Joyal semantics}
\frame[t]{\frametitle{The internal language of a topos}
  The internal language of a topos~$\E$ allows to
  \begin{enumerate}
    \item construct objects and morphisms of the topos,
    \item formulate statements about them and
    \item prove such statements
  \end{enumerate}
  in a \hil{naive element-based} language:

@@

  The \hil{little Zariski topos} of~$A$ is the category~$\Sh(\Spec(A))$ of
  set-valued sheaves on its spectrum. It contains a ring object~$A^\sim$
  with~$A^\sim(D(f)) = A[f^{-1}]$, called the \hil{structure sheaf}.

  \begin{block}{Motto (to be amended)}The structure sheaf~$A^\sim$ is a reification of
  all of the stalks~$A_\ppp$.\end{block}

@@

  \hil{Proof.} The structure sheaf~$A^\sim$ is the localisation
  \[ \underline{A}[\F^{-1}] \]
  of the constant sheaf~$\underline{A}$ at the \hil{generic filter}~$\F$, a
  sheaf with
  \[ \F_\ppp = A \setminus \ppp. \]
  The rings~$A$ and~$\underline{A}$ share all first-order
  properties.

  \hil{Remark.} Same statement for~$A$-modules~$M$ and their mirror
  image~$M^\sim = \underline{M}[\F^{-1}]$.
\end{frame}

@@

  \hil{Theorem.} If~$1 \neq 0$ in~$A$, there exists $f \neq 0$ in $A$ such that
  \begin{enumerate}
    \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
    \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$.
  \end{enumerate}
  \pause

@@

  \hil{Proof.} The claim is the translation of the fact that, internally
  in~$\Sh(\Spec(A))$, it is \hil{not not} the case that
  \begin{enumerate}
    \item $B^\sim$ and~$M^\sim$ are free over~$A^\sim$,
    \item $A^\sim \to B^\sim$ is of finite presentation, and
    \item $M^\sim$ is finitely presented as a module
    over~$B^\sim$.
  \end{enumerate}
\end{frame}

\begin{frame}
  Assume that~$B^\sim$ is generated by~$(x^i y^j)_{i,j\geq0}$ as an~$A^\sim$-module.
  It's \hil{not not} the case that either some generator can be expressed as a
  linear combination of others with smaller index, or not.

  \centering
  \begin{tikzpicture}[scale=0.4]
    \begin{scope}
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (6,0) -- (6,5) -- (5,5) -- (5,8);

      \fill[fill=red!40!white,opacity=0.5]
        (5,5) -- (6,5) -- (6,6) -- (5,6);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {1};
    \end{scope}

    \begin{scope}[xshift=9cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (5,8) -- (5,5) -- (8,5) -- (8,8);
      \draw
        (5,8) -- (5,5) -- (8,5);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (3,0) -- (3,5) -- (2,5) -- (2,8);

      \fill[fill=red!40!white,opacity=0.5]
        (2,5) -- (3,5) -- (3,6) -- (2,6);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {2};
    \end{scope}

    \begin{scope}[xshift=18cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (2,8) -- (2,5) -- (8,5) -- (8,8);
      \draw
        (2,8) -- (2,5) -- (8,5);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (7,0) -- (7,2) -- (6,2) -- (6,5) -- (2,5) -- (2,8);

      \fill[fill=red!40!white,opacity=0.5]
        (6,2) -- (7,2) -- (7,3) -- (6,3);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {3};
    \end{scope}

    \begin{scope}[yshift=-9cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (2,8) -- (2,5) -- (6,5) -- (6,2) -- (8,2) -- (8,8);
      \draw
        (2,8) -- (2,5) -- (6,5) -- (6,2) -- (8,2);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (6,0) -- (6,3) -- (5,3) -- (5,5) -- (2,5) -- (2,8);

      \fill[fill=red!40!white,opacity=0.5]
        (5,3) -- (6,3) -- (6,4) -- (5,4);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {4};
    \end{scope}

    \begin{scope}[yshift=-9cm,xshift=9cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (2,8) -- (2,5) -- (5,5) -- (5,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);

      \fill[fill=blue!40!white,opacity=0.5]
        (0,8) -- (0,0) -- (4,0) -- (4,3) -- (3,3) -- (3,5) -- (2,5) -- (2,8);

      \fill[fill=red!40!white,opacity=0.5]
        (3,3) -- (4,3) -- (4,4) -- (3,4);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {5};
    \end{scope}

    \begin{scope}[yshift=-9cm,xshift=18cm]
      \draw[step=1cm,gray,very thin] (0,0) grid (8,8);

      \draw
        (0,8) -- (0,0) -- (8,0);

      \fill[pattern=north east lines,pattern color=gray,opacity=0.5]
        (2,8) -- (2,5) -- (3,5) -- (3,3) -- (6,3) -- (6,2) -- (8,2) -- (8,8);
      \draw
        (2,8) -- (2,5) -- (3,5) -- (3,3) -- (6,3) -- (6,2) -- (8,2);

      \node[shape=circle,draw,inner sep=2pt] at (1,1) {6};
    \end{scope}
  \end{tikzpicture}
  \par
\end{frame}

\section{Outlook}

\begin{frame}{Outlook}
  \begin{itemize}
    \item Are there general guidelines for when using the internal perspective
    pays off?
    \item What about finer topologies (constructible, étale, fppf, \ldots)?
    \item How should synthetic algebraic geometry be extended?
  \end{itemize}

  {\vspace{-0.1em}\centering
  \rotatebox{90}{\tiny\scalebox{0.5}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
  \includegraphics[scale=0.27]{images/external-internal-small}
  \par\medskip\vspace{-0.5em}}

  \centering
  \hil{\href{https://tiny.cc/topos-notes}{https://tiny.cc/topos-notes}}
  \par
\end{frame}

@@

% Outlook (see research statement and mention algebraic geometry)
% Also mention synthetic algebraic geometry
% Include dictionary
